---
output: html_document
---
#Analysis of storm data for health and economic impact

**Synopsis**  
This is an analysis of storm data to study the impact on population health and economy. The data is sourced from U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database.

**Data Processing**  
The data is cleaned up after being read from the csv file. I have also use a manually created mapping file to standardize the event types because the CSV file contains miss-spelt types and lots on non-standard types. Here is the list of steps that summarizes the data processing done:
  
1. Read the csv.bz2 file
2. Trim the leading and trailing spaces from the EVTYPE field.
3. Write out a csv file with unique EVTYPE values
4. Manually map the EVTYPE values with the standard types from the [documentation PDF](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf). This is done to correct misspelt event types and non-standard types. 
5. Read the mapping CSV file created in the previous step.
6. Merge the mapped event types to the original data.
7. Prepare aggregated data for analysing the population health impact. This is simply a dataset with FATALITIES and INJURIES summed up for each event-type.
8. Prepare aggregated data for analysing the economic impact. This is simply a dataset with PROPDMG and CROPDMG summed up for each event-type. PROGDMGEXP and CROPEMGEXP field holds the magnitude of damage given in PROPDMG and CROPDMG fields. K means 'thousands', M 'millions' and B 'billions'. Other values are treated as 'ones'.
  
```{r echo=TRUE, cache=TRUE}
# setting working directory
setwd("c:/Users/karmacha/Documents/workshop/coursera/repdata-005/RepData_PeerAssessment2/")

# reading in the raw csv.bz2 file
rawdata <- read.csv(bzfile("repdata-data-StormData.csv.bz2"))
trim <- function(x) gsub("^\\s+|\\s+$", "", x)
rawdata$EVTYPE <- trim(toupper(rawdata$EVTYPE))

# writing out unique event type to file
# i intend to manually standardize the types so i get a nice category
# with less duplicates
write.csv(data.frame(sort(unique(rawdata$EVTYPE))), "storm_event_types_raw.csv")

# read the file with manually mapped event types
# the types listed in the guide pdf is used.
evtypemap <- read.csv("storm_event_types.txt")

# map the events to standardized types
cleandata <- merge(rawdata, evtypemap, by="EVTYPE")
```

```{r echo=TRUE}
### HEALTH IMPACT ###
# take only the data required for health impact
healthimpactdata <- cleandata[cleandata$FATALITIES > 0 | cleandata$INJURIES > 0,
                              c("STDTYPE", "FATALITIES", "INJURIES")]

# sum up casualties for each event type
healthimpactagg <- aggregate(cbind(FATALITIES, INJURIES)~STDTYPE,
                             data=healthimpactdata, FUN=sum)
healthimpactagg$total_casualties <- healthimpactagg$FATALITIES +
                                    healthimpactagg$INJURIES

# reorder data based on the impact for clear plot
healthimpactagg$STDTYPE <- factor(healthimpactagg$STDTYPE,
                                  levels=healthimpactagg[order(healthimpactagg$total_casualties), "STDTYPE"])

# keep only top 10
healthimpactagg <- head(healthimpactagg[order(-healthimpactagg$total_casualties), ], 10)

### ECONOMIC IMPACT ###
# take only the data required for economic impact
ecoimpactdata <- cleandata[cleandata$PROPDMG > 0 | cleandata$CROPDMG > 0,
                           c("STDTYPE", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]

# multiplication factor for the dollar values
# based on documentation K means 'thousands', M 'millions' and B 'billions'
# other characters are present too. I will treat other types as 'ones'.
pmfactor <- data.frame(EXP=c("", "K", "M", "B"), pfac=c(1, 1000, 1000000, 1000000000))
cmfactor <- data.frame(EXP=c("", "K", "M", "B"), cfac=c(1, 1000, 1000000, 1000000000))
ecoimpactdata$PROPDMGEXP <- toupper(ecoimpactdata$PROPDMGEXP)
ecoimpactdata$CROPDMGEXP <- toupper(ecoimpactdata$CROPDMGEXP)
ecoimpactdata <- merge(ecoimpactdata, pmfactor, by.x="PROPDMGEXP", by.y="EXP")
ecoimpactdata <- merge(ecoimpactdata, cmfactor, by.x="CROPDMGEXP", by.y="EXP")
ecoimpactdata$propdmgamt <- ecoimpactdata$PROPDMG * ecoimpactdata$pfac
ecoimpactdata$cropdmgamt <- ecoimpactdata$CROPDMG * ecoimpactdata$cfac
ecoimpactdata$totdmgamt <- ecoimpactdata$propdmgamt + ecoimpactdata$cropdmgamt

# sum up damages for each event type
ecoimpactagg <- aggregate(totdmgamt ~ STDTYPE, data=ecoimpactdata, FUN=sum)

# reorder data based on damage amount for clear plot
ecoimpactagg$STDTYPE <- factor(ecoimpactagg$STDTYPE,
                               levels=ecoimpactagg[order(ecoimpactagg$totdmgamt),
                                                   "STDTYPE"])
# keep only top 10
ecoimpactagg <- head(ecoimpactagg[order(-ecoimpactagg$totdmgamt), ], 10)

### YEARLY TIMESERIES data ###
tsdata <- cleandata[cleandata$PROPDMG > 0 | cleandata$CROPDMG > 0,
                           c("BGN_DATE", "STDTYPE", "FATALITIES", "INJURIES",
                             "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
tsdata$PROPDMGEXP <- toupper(tsdata$PROPDMGEXP)
tsdata$CROPDMGEXP <- toupper(tsdata$CROPDMGEXP)
tsdata$date <- as.Date(tsdata$BGN_DATE, format = "%m/%d/%Y")
tsdata <- merge(tsdata, pmfactor, by.x="PROPDMGEXP", by.y="EXP")
tsdata <- merge(tsdata, cmfactor, by.x="CROPDMGEXP", by.y="EXP")
tsdata$propdmgamt <- tsdata$PROPDMG * tsdata$pfac
tsdata$cropdmgamt <- tsdata$CROPDMG * tsdata$cfac
tsagg <- aggregate(cbind(FATALITIES, INJURIES, propdmgamt, cropdmgamt)~date+STDTYPE,
                   data=tsdata,
                   FUN=sum)
library(reshape)
tsaggm <- melt(tsagg, id=c("date", "STDTYPE"))
```

**Results**  
Here is a plot of the total number of casualties for each event type.
```{r fig.height=3}
library(ggplot2)
g1 <- ggplot(healthimpactagg)
g1 + geom_bar(aes(x=STDTYPE, y=total_casualties), stat="identity") + 
    xlab("Event Type") +
    ylab("Total Casualties (Fatalities + Injuries)") +
    ggtitle("Top 10 events based on health impact") +
    coord_flip()
```
  
It is evident that Tornado has caused the highest number of casualties. Heat, thunderstorm wind, flash flood and lightning are other events with significant impact on the population health.
  
  
Up next is a plot of total economic damage in billions against the event types.
  
```{r fig.height=3}
g2 <- ggplot(ecoimpactagg)
g2 + geom_bar(aes(x=STDTYPE, y=totdmgamt/1000000000), stat="identity") +
    xlab("Event Type") +
    ylab("Total Damages in billions (Property + Crop)") +
    ggtitle("Top 10 events based on economic impact") +
    coord_flip()
```
  
Flash flood seems to have the greatest impact on economic damage. Hurricane(Typhoon), tornado and high-surf have big impacts too.

